---
title: "GDP components over time and among countries"
date: '2020-10-20'
description: Comparing countries by GDP components over time
output:
  pdf_document: default
  html_document:
    df_print: paged
image: ''
keywords: ''
draft: no
slug: project4
categories:
- ''
- ''
---

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width= 7, 
  fig.height=7,
  fig.align = "center"
)
```



```{r load-libraries, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
library(tidyquant)
library(scales)
library(car)

```

There are four main components of gross domestic product: personal consumption (C), business investment (I), government spending (G) and net exports (exports - imports). Here you can read more about GDP and the different approaches in calculating at the [Wikipedia GDP page](https://en.wikipedia.org/wiki/Gross_domestic_product).

I will consider the GDP data from the [United Nations' National Accounts Main Aggregates Database](https://unstats.un.org/unsd/snaama/Downloads), which contains estimates of total GDP and its components for all countries from 1970 to today.

We will see how different components of GDP changed over time and compare different countries  by its GDP components estimates.

First, let us upload the data.

```{r read_GDP_data}
library(readxl)
UN_GDP_data  <- read_excel(here::here("data", "Download-GDPconstant-USD-countries.xls"), # Excel filename
                sheet="Download-GDPconstant-USD-countr", # Sheet name
                skip=2) # Number of rows to skip

```

Then we need to tidy the data and transform it into long format

```{r reshape_GDP_data}

#pivot_longer to transform data from wide format to long
tidy_GDP_data <- UN_GDP_data %>% pivot_longer(-c(CountryID, Country, IndicatorName), #which columns to preserve
                                               values_to = "Value", #the name of values column
                                               names_to = "year") %>%  #column with names of previous columns with years
                                 filter(!is.na(Value)) %>% #remain only finite values 
                                 mutate(bln_value = Value/10^9, #divide by 1 bln to get bln values
                                        year = as.numeric(year)) #transform year from chr to numeric
#see columns type of df
glimpse(tidy_GDP_data)

# Let us compare GDP components for these 3 countries
country_list <- c("United States","India", "Germany")

```

Now I would like to construct a plot that compares different countries by GDP components.

```{r gdp1, out.width="100%", fig.width= 8, fig.height=5.5}

options(scipen = 999)

#recode is used to rename text rows: first argument is the column where we want to change names, then all replacements in rows
tidy_GDP_data$IndicatorName <-  car::recode(tidy_GDP_data$IndicatorName, 
                                       "'Exports of goods and services' = 'Exports';
                                       'Imports of goods and services' = 'Imports';
                                       'General government final consumption expenditure' = 'Government expenditure';
                                       'Household consumption expenditure (including Non-profit institutions serving households)' = 'Household expenditure'")

gdp_components <- c("Gross capital formation", "Exports", "Government expenditure", "Household expenditure", "Imports")

gdp_data_filtered <- tidy_GDP_data %>% 
                      filter(Country %in% country_list & #filter data
                             IndicatorName %in% gdp_components) %>% 
                      mutate(IndicatorName = factor(IndicatorName, levels = gdp_components, #create labels for gdp components
                                                                    labels = gdp_components))
  
ggplot(gdp_data_filtered,
       aes(x = year, y = bln_value, color = IndicatorName)) + #main set up
  geom_line(aes(group = IndicatorName)) + #add lines by indicator names
  facet_wrap(~Country) + #add tiles by country
  scale_x_continuous(breaks = seq(1970, 2020, by = 10)) + #make scale from 1970 to 2020 by step = 10
  theme_bw() + #specify white theme
  labs(title = "GDP Components over time", #provide names for title, subtitle, x,y
       subtitle = "In constant 2010 USD",
       x = "",
       y = "Billion US$",
       colour = "Components of GDP") + #change the legend name
  theme(plot.title = element_text(face = "bold"), #make title bold
        panel.grid.minor = element_blank()) #get rid of minor lines in the background


```


Secondly, recall that GDP is the sum of Household Expenditure (Consumption *C*), Gross Capital Formation (business investment *I*), Government Expenditure (G) and Net Exports (exports - imports). Even though there is an indicator `Gross Domestic Product (GDP)` in your dataframe, I would like you to calculate it given its components discussed above.

However, I have some doubts that GDP variable in the data is calculated correctly. Below in the code we calculate GDP as the sum of Household Expenditure (Consumption *C*), Gross Capital Formation (business investment *I*), Government Expenditure (G) and Net Exports (exports - imports). Let's answer the question: What is the percentage difference between what we calculated as GDP and the GDP figure included in the dataframe?

```{r perc_diff, out.width="100%", fig.width= 8, fig.height=6}

gdp_percentage_diff <- tidy_GDP_data %>% 
                       pivot_wider(names_from = IndicatorName, 
                                   values_from = bln_value,
                                   id_cols = c(Country, year)) %>% #transform groups from one column to multiple columns
                       mutate(GDP_calculated = `Gross capital formation` + 
                                               `Exports` + 
                                               `Government expenditure` +
                                               `Household expenditure` - 
                                               `Imports`,
                              GDP_percentage_difference = GDP_calculated / `Gross Domestic Product (GDP)` - 1) %>% 
                      group_by(Country) %>% 
                      filter(!is.na(GDP_percentage_difference)) %>% 
                      summarise(mean_diff = abs(mean(GDP_percentage_difference, na.rm = T) * 100)) %>% 
                      arrange(desc(mean_diff)) #descending sort by mean difference %

gdp_percentage_diff %>% 
  head(n = 25) %>% #select top 25 countries (sorted previously by arrange)
ggplot(aes(y = reorder(Country, mean_diff), #sort countries by mean difference
           x = mean_diff)) +
geom_col(fill = "#00B81F") +
theme_bw() +
labs(x = "Country", 
     y = "Absolute mean difference", 
     title = "Yemen has problems with calculating GDP",
     subtitle = "Top 25 countries that have the highest mean absolute difference \n between calculated GDP and GDP figure ")


#average mean by all countries
mean(abs(gdp_percentage_diff$mean_diff))

```

Yemen has the highest percentage difference between calculated GDP and provided GDP in the figure and deviates by 34% from the figure. Moreover, other countries in this list are  located either in the islands, in the Middle East or in Africa.

The average absolute mean difference percentage among all countries is 14.22% which is quite high. However, our top-25 contributes a lot to this  figure.

Furthermore, I suggest to calculate GDP components as a percentage of total GDP and compare them by countries.

```{r gdp2, out.width="100%", fig.width= 8, fig.height=6}

gdp_data_filtered2 <- gdp_data_filtered %>% 
                      pivot_wider(names_from = IndicatorName, 
                                  values_from = bln_value,
                                  id_cols = c(Country, year)) %>% 
                      mutate(`Net Exports` = Exports - Imports,
                             Imports = - Imports) %>% #we used 2 mutates because of structure of dplyr: it calcs simultaneously
                      mutate(GDP = rowSums(.[3:7]),
                             `Household expenditure` = `Household expenditure` / GDP,
                             `Government expenditure` = `Government expenditure` / GDP,
                             `Gross capital formation` = `Gross capital formation` / GDP,
                             `Net Exports` = `Net Exports` / GDP) %>% 
                      select(-c(Exports, Imports, GDP)) %>% 
                      pivot_longer(-c(Country, year), #which columns to preserve
                                   values_to = "percentage", #the name of values column
                                   names_to = "IndicatorName")

ggplot(gdp_data_filtered2,
       aes(x = year, y = percentage, color = IndicatorName)) + #main set up
  geom_line(aes(group = IndicatorName)) + #add lines by indicator names
  facet_wrap(~Country) + #add tiles by country
  scale_x_continuous(breaks = seq(1970, 2020, by = 10)) + #make scale from 1970 to 2020 by step = 10
  theme_bw() + #specify white theme
  labs(title = "GDP and its breakdown at constant 2010 prices in US Dollars", #provide names for title, subtitle, x,y
       caption = "Source: United Nations, https://unstats.un.org/unsd/snaama/Downloads",
       x = "",
       y = "proportion",
       colour = "") + #change the legend name
  theme(panel.grid.minor = element_blank()) #get rid of minor lines in the background
  

```

First of all, Germany and United States are developed countries which means that they should have similar patterns but some components of GDP demonstrate otherwise, while India is an emerging country. In our developed countries investment and government components of GDP intersected several times while in India there was a huge difference between them in 1970s and has increased during the 50-year period. The investment share in both developed countries is at the same level - 20% of total GDP - while in India it has almost doubled during the period.

Household expenditure in Germany and India has decreased during the discussed period but in America there was a reverse situation - consumer spending proportion has risen. Furthermore, household expenditure is still the largest part of GDP in all three countries.

Also, it is very interesting to consider Net exports proportion. India and USA has the similar movement pattern in this component: it moves in one direction in both countries and also is negative which means that imports outweigh exports. In Germany net exports proportion was around 0% but since 2000 it has increased almost to 10% of GDP.

Finally, let's construct function that will help us to construct the graphs for any countries that we use as input.

```{r gdp_general, out.width="100%", fig.width= 7, fig.height=6}
library(gridExtra)
#I would like to speculate this task and advance further. 
#I will create a function that can draw the graph with any countries I would like to compare.

UN_GDP_data  <- read_excel(here::here("data", "Download-GDPconstant-USD-countries.xls"), # Excel filename
                sheet="Download-GDPconstant-USD-countr", # Sheet name
                skip=2) # Number of rows to skip

# create a function where you plug in 
compare_countries_by_gdp <- function(UN_GDP_data, country_list) {
  

#pivot_longer to transform data from wide format to long
tidy_GDP_data <- UN_GDP_data %>% pivot_longer(-c(CountryID, Country, IndicatorName), #which columns to preserve
                                               values_to = "Value", #the name of values column
                                               names_to = "year") %>%  #column with names of previous columns with years
                                 filter(!is.na(Value)) %>% #remain only finite values 
                                 mutate(bln_value = Value/10^9, #divide by 1 bln to get bln values
                                        year = as.numeric(year)) #transform year from chr to numeric

  #recode is used to rename text rows: first argument is the column where we want to change names, then all replacements in rows
tidy_GDP_data$IndicatorName <-  car::recode(tidy_GDP_data$IndicatorName, 
                                       "'Exports of goods and services' = 'Exports';
                                       'Imports of goods and services' = 'Imports';
                                       'General government final consumption expenditure' = 'Government expenditure';
                                       'Household consumption expenditure (including Non-profit institutions serving households)' = 'Household expenditure'")

gdp_components <- c("Gross capital formation", "Exports", "Government expenditure", "Household expenditure", "Imports")

gdp_data_filtered <- tidy_GDP_data %>% 
                      filter(Country %in% country_list & #filter data
                             IndicatorName %in% gdp_components) %>% 
                      mutate(IndicatorName = factor(IndicatorName, levels = gdp_components, #create labels for gdp components
                                                                    labels = gdp_components))
gdp_value <- ggplot(gdp_data_filtered,
                   aes(x = year, y = bln_value, color = IndicatorName)) + #main set up
              geom_line(aes(group = IndicatorName)) + #add lines by indicator names
              facet_wrap(~Country) + #add tiles by country
              scale_x_continuous(breaks = seq(1970, 2020, by = 10)) + #make scale from 1970 to 2020 by step = 10
              theme_bw() + #specify white theme
              labs(title = "GDP Components over time", #provide names for title, subtitle, x,y
                   subtitle = "In constant 2010 USD",
                   x = "",
                   y = "Billion US$",
                   colour = "Components of GDP") + #change the legend name
              theme(plot.title = element_text(face = "bold"), #make title bold
                    panel.grid.minor = element_blank()) #get rid of minor lines in the background



gdp_data_filtered2 <- gdp_data_filtered %>% 
                      pivot_wider(names_from = IndicatorName, 
                                  values_from = bln_value,
                                  id_cols = c(Country, year)) %>% 
                      mutate(`Net Exports` = Exports - Imports,
                             Imports = - Imports) %>% #we used 2 mutates because of structure of dplyr: it calcs simultaneously
                      mutate(GDP = rowSums(.[3:7]),
                             `Household expenditure` = `Household expenditure` / GDP,
                             `Government expenditure` = `Government expenditure` / GDP,
                             `Gross capital formation` = `Gross capital formation` / GDP,
                             `Net Exports` = `Net Exports` / GDP) %>% 
                      select(-c(Exports, Imports, GDP)) %>% 
                      pivot_longer(-c(Country, year), #which columns to preserve
                                   values_to = "percentage", #the name of values column
                                   names_to = "IndicatorName")

gdp_prop <- ggplot(gdp_data_filtered2,
                   aes(x = year, y = percentage, color = IndicatorName)) + #main set up
              geom_line(aes(group = IndicatorName)) + #add lines by indicator names
              facet_wrap(~Country) + #add tiles by country
              scale_x_continuous(breaks = seq(1970, 2020, by = 10)) + #make scale from 1970 to 2020 by step = 10
              theme_bw() + #specify white theme
              labs(title = "GDP and its breakdown at constant 2010 prices in US Dollars", #provide names for title, subtitle, x,y
                   caption = "Source: United Nations, https://unstats.un.org/unsd/snaama/Downloads",
                   x = "",
                   y = "proportion",
                   colour = "") + #change the legend name
              theme(panel.grid.minor = element_blank()) #get rid of minor lines in the background


return(grid.arrange(gdp_value, gdp_prop, nrow = 2))
}
```

Now let run the function and compare Russian Federation, France, Greece, and Germany by its GDP components.

```{r graphs}
country_list <- c("Russian Federation", "France", "Greece", "Germany")
compare_countries_by_gdp(UN_GDP_data, country_list)

```

Overall, we compared countries by its GDP components and learned how to write a function that compares any number of countries by its GDP over time.


